---
title: "Dates and Times with lubridate"
output:
  pdf_document:
    toc: yes
  html_document:
    highlight: zenburn
    theme: cosmo
    toc: yes
---

###Guaranteed to work with an "en_US.UTF-8"locale
```{r}
Sys.getlocale("LC_TIME")
library(lubridate)
```
### lubridate contains many useful functions
```
help(package = lubridate)
```
###today() function returns today's date
```{r}
this_day <- today()
this_day
```
###There are three components to this date. In order, they are year, month, and day.
```{r}
year(this_day)
month(this_day)
day(this_day)
```
###Get the day of the week from this_day using the wday() function
```{r}
wday(this_day)
wday(this_day, label = TRUE)
```
###The now() function returns the date-time representing this exact moment in time.
```{r}
now()
this_moment <- now()
this_moment
```
> Just like with dates, we can extract the year, month, day, or day of week. However, we can also use hour(), minute(), and second() to extract specific time information.

```{r}
hour(this_moment)
minute(this_moment)
second(this_moment)
```
> lubridate offers a variety of functions for parsing date-times. These functions take the form of ymd(), dmy(), hms(), ymd_hms(), etc., where each letter in the name of the function stands for the location of years (y), months (m), days (d), hours (h), minutes (m), and/or seconds (s) in the date-time being read in.

```{r}
my_date <- ymd("1989-05-17")
my_date
class(my_date)
ymd("1989 May 17")
mdy("March 12, 1975")
dmy( 25081985)
ymd("192012")
ymd("1920/1/2")
```
### In addition to dates, lubridate can parse date-times.
```
> dt1
[1] "2014-08-23 17:23:02"
ymd_hms(dt1)
[1] "2014-08-23 17:23:02 UTC"
```
```{r}
hms("03:22:14")
```
###lubridate is also capable of handling vectors of dates, which is particularly helpful when you need to parse an entire column of data.
```
> dt2
[1] "2014-05-14" "2014-09-22" "2014-07-11"

> ymd(dt2)
[1] "2014-05-14 UTC" "2014-09-22 UTC" "2014-07-11 UTC"
```
### The update() function allows us to update one or more components of a date-time.
```{r}
update(this_moment, hours = 8, minutes = 34, seconds = 55)
this_moment
```
###To find the current date in New York
```{r}
nyc <- now("America/New_York")
nyc
#flight is the day after tomorrow (in New York time)
nyc + days(2)
depart <- nyc + days(2)
depart
depart <- update(depart, hours = 17, minutes = 34)
depart
```
###what time will flight arrive the airport in Hong Kong?
```{r}
arrive <- depart + hours(15) + minutes(50)
```
###The with_tz() function returns a date-time as it would appear in another time zone
```{r}
arrive <- with_tz(arrive,"Asia/Hong_Kong")
arrive
```
### how long it has been from June 17, 2008 in Singapore
```{r}
last_time <- mdy("June 17, 2008", tz = "Singapore")
last_time
how_long <- new_interval(last_time, arrive)
as.period(how_long)
```




