---
title: "Grouping and Chaining with dplyr"
output:
  pdf_document:
    toc: yes
  html_document:
    highlight: zenburn
    theme: cosmo
    toc: yes
---
> There are five main data manipulation 'verbs' in dplyr: select(), filter(), arrange(), mutate(), and summarize().  
The last of these, summarize(), is most powerful when applied to grouped data.  
The main idea behind grouping data is that you want to break up your dataset into groups of rows based on the values of one or more variables. The group_by() function is reponsible for doing this.
###Initialization
```{r}
library(dplyr)
path2csv <- "/Library/Frameworks/R.framework/Versions/3.1/Resources/library/swirl/Courses/Getting_and_Cleaning_Data/Manipulating_Data_with_dplyr/2014-07-08.csv"
path2csv %>%
    read.csv( stringsAsFactors =FALSE) %>%
    tbl_df() -> cran
rm("mydf")
cran
```

##gtoup_by( )
```{r}
?group_by
```
###Group cran by the package variable and store the result in a new variable called by_package
```{r}
by_package <- group_by(cran, package)
by_package
```
> Applied mean(size) to the original tbl_df via summarize(), it returned a single number -- the mean of all values in the size column.   

### Wouldn't it be so much more interesting to look at the mean download size for each unique package?  
### That's exactly what you'll get if you use summarize() to apply mean(size) to the grouped data in by_package.
```{r}
summarize(by_package, mean(size))
```
### Compute four values, in the following order, from the grouped data: count = n(), unique = n_distinct(ip_id), countries = n_distinct(country), avg_bytes = mean(size)
```{r}
pack_sum <- summarize(by_package,
                      count = n(),
                      unique = n_distinct(ip_id),
                      countries = n_distinct(country),
                      avg_bytes = mean(size))
pack_sum
```
The 'count' column, created with n(), contains the total number of rows (i.e. downloads) for each package. The 'unique' column, created with n_distinct(ip_id), gives the total number of unique downloads for each package, as measure by the number of distinct ip_id's. The 'countries' column, created with n_distinct(country), provides the number of countries in which the each package was downloaded. And finally, the 'avg_bytes' column, created with mean(size), contains the mean download size (in bytes) for each package.
###splits the data into the top 1% and bottom 99% of packages based on total downloads
```{r}
quantile(pack_sum$count, probs = 0.99)
```
###Isolate only those packages which had more than 679 total downloads
```{r}
top_counts <- filter(pack_sum, count > 679)
top_counts
```
> By default, if the dataset your working with has more than 100 rows, dplyr will only print the first 10. Otherwise, it will print all rows, since it's not as overwhelming.

### See the first 20
```{r}
head(top_counts, 20)
```
###Arrange() the rows of top_counts based on the 'count' column
```{r}
arrange(top_counts, desc(count))
```
###Find the 0.99, or 99%, quantile for the 'unique' variable
```{r}
quantile(pack_sum$unique, probs = 0.99)
```
### Select all rows corresponding to values of 'unique' that are strictly greater than 465
```{r}
top_unique <- filter(pack_sum, unique >465)
top_unique
```
### Arrange() top_unique by the 'unique' column, in descending order
```{r}
arrange(top_unique, desc(unique))
```
##chaing(piping)
> Chaining allows you to string together multiple function calls in a way that is compact and readable, while still accomplishing the desired result.

```
result3 <-
  cran %>%
  group_by(package) %>%
  summarize(count = n(),
            unique = n_distinct(ip_id),
            countries = n_distinct(country),
            avg_bytes = mean(size)
  ) %>%
  filter(countries > 60) %>%
  arrange(desc(countries), avg_bytes)
print(result3)
```
### Select() the following columns from cran, ip_id, country, package, size
```{r}
cran %>%
  select(ip_id, country, package, size) %>%
  #add a column called size_mb that contains
    mutate(size_mb = size / 2^20) %>%
  #select all rows for which size_mb is less than or equal to (<=) 0.5.
      filter(size_mb <= 0.5) %>%
  #arrange() the result by size_mb, in descending order.
        arrange(desc(size_mb)) %>%
          print
```

