---
title: "Datacamp 1"
output: word_document
---
##Load the monthly Starbucks return data
###Load Data from Web
```{r}
data_url = "http://faculty.washington.edu/ezivot/econ424/sbuxPrices.csv"

sbux_df = read.csv(file=data_url,header=TRUE,sep=",",stringsAsFactors =FALSE)

```
    
##Get a feel for the data
```{r}
str(sbux_df)
head(sbux_df)
tail(sbux_df)
class(sbux_df$Date)
```
   
   
##Extract the price data
```{r}
closing_prices = sbux_df[,"Adj.Close",drop=FALSE]

head(closing_prices)
```
   
   
##Find indices associated with the dates 3/1/1994 and 3/1/1995
### Find indices associated with the dates 3/1/1994 and 3/1/1995
```{r}
index_1 = which(sbux_df$Date=="3/1/1994")

index_2 = which(sbux_df$Date=="3/1/1995")

some_prices = sbux_df[index_1:index_2,"Adj.Close"]

some_prices
```
   
   
##Subset directly on dates
### Create a new data frame that contains the price data with the dates as the row names
```{r}
sbux_prices_df = sbux_df[, "Adj.Close", drop=FALSE]
rownames(sbux_prices_df) = sbux_df$Date
head(sbux_prices_df)
```
### With Dates as rownames, you can subset directly on the dates.
### Find indices associated with the dates 3/1/1994 and 3/1/1995.
```{r}
price_1 = sbux_prices_df["3/1/1994",1]
price_2 = sbux_prices_df["3/1/1995",1]
price_1;price_2
```
   
   
##Plot the price data
```{r}
plot(sbux_df$Adj.Close,type="l",col="blue",lwd=2,ylab="Adjusted close",main="Monthly closing price of SBUX")
```
   
   
##Calculate simple returns
```{r}
sbux_prices_df = sbux_df[, "Adj.Close", drop=FALSE]
```
### Denote n the number of time periods
```{r}
n = nrow(sbux_prices_df)
sbux_ret = (sbux_prices_df[2:n,1]-sbux_prices_df[1:(n-1),1])/sbux_prices_df[1:(n-1),1]
```
### Notice that sbux_ret is not a data frame object
```{r}
class(sbux_ret)
```
   
   
##Add dates to simple return vector
### Now add dates as names to the vector and print the first elements of sbux_ret to the console to check
```{r}
names(sbux_ret)=sbux_df$Date[2:n]
head(sbux_ret)
```
   
   
## Compute continuously compounded 1-month returns
```{r}
sbux_ccret = log(sbux_prices_df[2:n, 1])- log(sbux_prices_df[1:(n-1), 1])
```
### Assign names to the continuously compounded 1-month returns
```{r}
names(sbux_ccret) = sbux_df$Date[2:n]
```
### Show sbux_ccret
```{r}
head(sbux_ccret)
```
   
   
## Compare the simple and cc returns
```{r}
head(cbind(sbux_ret,sbux_ccret))
```
   
   
##Graphically compare the simple and continuously compounded returns
### Plot the returns on the same graph
```{r}
# Plot the returns on the same graph
plot(sbux_ret, type="l", col="blue", lwd=2, ylab="Return",
               main="Monthly Returns on SBUX")
# Add horizontal line at zero
abline(h=0)
# Add a legend
legend(x="bottomright", legend=c("Simple", "CC"), 
       lty=1, lwd=2, col=c("blue","red"))
# Add the continuously compounded returns
lines(sbux_ccret,col="red",lwd=2)
```
   
   
##Calculate growth of $1 invested in SBUX
### Compute gross returns
```{r}
sbux_gret = sbux_ret+1
```
### Compute future values
```{r}
sbux_fv = cumprod(sbux_gret)
```

### Plot the evolution of the $1 invested in SBUX as a function of time
```{r}
plot(sbux_fv, type="l", col="blue", lwd=2, ylab="Dollars", 
     main="FV of $1 invested in SBUX")
```
