---
title: "Tidying Data with tidyr"
output:
  pdf_document:
    toc: yes
  html_document:
    highlight: zenburn
    theme: united
    toc: yes
---
```{r}
library(tidyr)
```
The author of tidyr, Hadley Wickham, discusses his philosophy of tidy data in his 'Tidy Data' paper: ![http://vita.had.co.nz/papers/tidy-data.pdf]
> Tidy data is formatted in a standard way that facilitates exploration and analysis and works
seemlessly with other tidy data tools. Specifically, tidy data satisfies three conditions:  
1) Each variable forms a column  
2) Each observation forms a row  
3) Each type of observational unit forms a table  

> Any dataset that doesn't satisfy these conditions is considered 'messy' data. Therefore, all of the following are characteristics of messy data, EXCEPT...  
1: Column headers are values, not variable names  
2: Variables are stored in both rows and columns  
3: A single observational unit is stored in multiple tables  
4: Multiple types of observational units are stored in the same table  
5: Multiple variables are stored in one column  

### The first problem is when you have column headers that are values, not variable names.
```
  grade male female
1     A    1      5
2     B    5      0
3     C    5      2
4     D    5      5
5     E    7      4
```
This dataset actually has three variables: grade, sex, and count. The first variable, grade, is already a column, so that should remain as it is. The second variable, sex, is captured by the second and third column headings. The third variable, count, is the number of students for each combination of grade and sex.
##gather( )
```{r, eval=FALSE}
?gather
```
### To tidy the students data, need to have one column for each of these three variables
```
> gather(students, sex, count, -grade)
   grade    sex count
1      A   male     1
2      B   male     5
3      C   male     5
4      D   male     5
5      E   male     7
6      A female     5
7      B female     0
8      C female     2
9      D female     5
10     E female     4
```
Each row of the data now represents exactly one observation, characterized by a unique combination of the grade and sex variables.

###when multiple variables are stored in one column
```
> students2
  grade male_1 female_1 male_2 female_2
1     A      3        4      3        4
2     B      6        4      3        5
3     C      7        4      3        8
4     D      4        0      8        1
5     E      1        1      2        7
```
students2 suffers from the same messy data problem of having column headers that are values (male_1, female_1, etc.) and not variable names (sex, class, and count). it also has multiple variables stored in each column (sex and class), which is another common symptom of messy data.
###using gather() to stack the columns of students2
```
> res <-gather(students2, sex_class, count, -grade)
> res
   grade sex_class count
1      A    male_1     3
2      B    male_1     6
3      C    male_1     7
4      D    male_1     4
5      E    male_1     1
6      A  female_1     4
7      B  female_1     4
8      C  female_1     4
9      D  female_1     0
10     E  female_1     1
11     A    male_2     3
12     B    male_2     3
13     C    male_2     3
14     D    male_2     8
15     E    male_2     2
16     A  female_2     4
17     B  female_2     5
18     C  female_2     8
19     D  female_2     1
20     E  female_2     7

```
Two different variables, sex and class, are stored together in the sex_class column.

##separate( )
###Call separate() on res to split the sex_class column into sex and class
```
> separate(res, col = sex_class, into = c("sex", "class"))
   grade    sex class count
1      A   male     1     3
2      B   male     1     6
3      C   male     1     7
4      D   male     1     4
5      E   male     1     1
6      A female     1     4
7      B female     1     4
8      C female     1     4
9      D female     1     0
10     E female     1     1
11     A   male     2     3
12     B   male     2     3
13     C   male     2     3
14     D   male     2     8
15     E   male     2     2
16     A female     2     4
17     B female     2     5
18     C female     2     8
19     D female     2     1
20     E female     2     7
```
separate() was able to figure out on its own how to separate the sex_class column. Unless you request otherwise with the 'sep' argument, it splits on non-alphanumeric values.
In other words, it assumes that the values are separated by something other than a letter or number (in this case, an underscore.)

### Tidying students2 required both gather() and separate(), need us to save an intermediate result (res). However, just like with dplyr, you can use the %>% operator to chain multiple function calls together.
```
students2 %>%
  gather( sex_class, count, -grade ) %>%
  separate( col = sex_class, into = c("sex", "class")) %>%
  print
```
###A third symptom of messy data is when variables are stored in both rows and columns.
```
> students3
    name    test class1 class2 class3 class4 class5
1  Sally midterm      A   <NA>      B   <NA>   <NA>
2  Sally   final      C   <NA>      C   <NA>   <NA>
3   Jeff midterm   <NA>      D   <NA>      A   <NA>
4   Jeff   final   <NA>      E   <NA>      C   <NA>
5  Roger midterm   <NA>      C   <NA>   <NA>      B
6  Roger   final   <NA>      A   <NA>   <NA>      A
7  Karen midterm   <NA>   <NA>      C      A   <NA>
8  Karen   final   <NA>   <NA>      C      A   <NA>
9  Brian midterm      B   <NA>   <NA>   <NA>      A
10 Brian   final      B   <NA>   <NA>   <NA>      C
```
Call gather() to gather the columns class1 through through class5 into a new variable called class. The 'key' should be class, and the 'value' should be grade.   
tidyr makes it easy to reference multiple adjacent columns with class1:class5, just like with sequences of numbers.
```
students3 %>%
  gather( class, grade, class1:class5 , na.rm= TRUE) %>%
  print

    name    test  class grade
1  Sally midterm class1     A
2  Sally   final class1     C
3  Brian midterm class1     B
4  Brian   final class1     B
5   Jeff midterm class2     D
6   Jeff   final class2     E
7  Roger midterm class2     C
8  Roger   final class2     A
9  Sally midterm class3     B
10 Sally   final class3     C
11 Karen midterm class3     C
12 Karen   final class3     C
13  Jeff midterm class4     A
14  Jeff   final class4     C
15 Karen midterm class4     A
16 Karen   final class4     A
17 Roger midterm class5     B
18 Roger   final class5     A
19 Brian midterm class5     A
20 Brian   final class5     C
```
##spread( )
###Turn the values of the test column, midterm and final, into column headers (i.e. variables).
```
students3 %>%
  gather(class, grade, class1:class5, na.rm = TRUE) %>%
  spread( test, grade) %>%
  print
  
    name  class final midterm
1  Brian class1     B       B
2  Brian class5     C       A
3   Jeff class2     E       D
4   Jeff class4     C       A
5  Karen class3     C       C
6  Karen class4     A       A
7  Roger class2     A       C
8  Roger class5     A       B
9  Sally class1     C       A
10 Sally class3     C       B
```
##extract_numeric( )
###Let the values in the class columns to be 1, 2, ..., 5 and not class1, class2, ..., class5
###Use the mutate() function from dplyr along with extract_numeric().
```
students3 %>%
  gather(class, grade, class1:class5, na.rm = TRUE) %>%
  spread(test, grade) %>%
  ### Call to mutate() goes here %>%
  mutate(class = extract_numeric(class)) %>%
  print

    name class final midterm
1  Brian     1     B       B
2  Brian     5     C       A
3   Jeff     2     E       D
4   Jeff     4     C       A
5  Karen     3     C       C
6  Karen     4     A       A
7  Roger     2     A       C
8  Roger     5     A       B
9  Sally     1     C       A
10 Sally     3     C       B
```
###multiple observational units are stored in the same table
```
> students4
    id  name sex class midterm final
1  168 Brian   F     1       B     B
2  168 Brian   F     5       A     C
3  588 Sally   M     1       A     C
4  588 Sally   M     3       B     C
5  710  Jeff   M     2       D     E
6  710  Jeff   M     4       A     C
7  731 Roger   F     2       C     A
8  731 Roger   F     5       B     A
9  908 Karen   M     3       C     C
10 908 Karen   M     4       A     A
```
notice that each id, name, and sex is repeated twice, which seems quite redundant.
###selecting the id, name, and sex column from students4 and storing the result in student_info.
```
student_info <- students4 %>%
  select(id, name, sex) %>%
  remove duplicate rows from student_info.
  unique %>%
  print

   id  name sex
1 168 Brian   F
3 588 Sally   M
5 710  Jeff   M
7 731 Roger   F
9 908 Karen   M
```
### Select() the id, class, midterm, and final columns (in that order) and store the result in gradebook.
```
gradebook <- students4 %>%
  select(id, class, midterm, final) %>%
  print
  
    id class midterm final
1  168     1       B     B
2  168     5       A     C
3  588     1       A     C
4  588     3       B     C
5  710     2       D     E
6  710     4       A     C
7  731     2       C     A
8  731     5       B     A
9  908     3       C     C
10 908     4       A     A
```
> It's important to note that we left the id column in both tables. In the world of relational databases, 'id' is called our 'primary key' since it allows us to connect each student listed in student_info with their grades listed in gradebook. Without a unique identifier, we might not know how the tables are related. (In this case, we could have also used the name variable, since each student happens to have a unique name.)

```
> passed
   name class final
1 Brian     1     B
2 Roger     2     A
3 Roger     5     A
4 Karen     4     A

> failed
   name class final
1 Brian     5     C
2 Sally     1     C
3 Sally     3     C
4  Jeff     2     E
5  Jeff     4     C
6 Karen     3     C

passed <- passed %>%mutate( status="passed")
failed <- failed %>%mutate( status="failed")

> passed
   name class final status
1 Brian     1     B passed
2 Roger     2     A passed
3 Roger     5     A passed
4 Karen     4     A passed

> failed
   name class final status
1 Brian     5     C failed
2 Sally     1     C failed
3 Sally     3     C failed
4  Jeff     2     E failed
5  Jeff     4     C failed
6 Karen     3     C failed

rbind_list(passed, failed)
    name class final status
1  Brian     1     B passed
2  Roger     2     A passed
3  Roger     5     A passed
4  Karen     4     A passed
5  Brian     5     C failed
6  Sally     1     C failed
7  Sally     3     C failed
8   Jeff     2     E failed
9   Jeff     4     C failed
10 Karen     3     C failed
```
> Of course, we could arrange the rows however we wish at this point, but the important thing is that each row is an observation, each column is a variable, and the table contains a single observational unit. Thus, the data are tidy.

##SAT EXAMPLE
> The SAT is a popular college-readiness exam in the United States that consists of three sections: critical reading, mathematics, and writing. Students can earn up to 800 points on each section. This dataset presents the total number of students, for each combination of exam section and sex, within each of six score ranges. 

```
> sat
Source: local data frame [6 x 10]

  score_range read_male read_fem read_total math_male math_fem math_total write_male write_fem
1     700–800     40151    38898      79049     74461    46040     120501      31574     39101
2     600-690    121950   126084     248034    162564   133954     296518     100963    125368
3     500-590    227141   259553     486694    233141   257678     490819     202326    247239
4     400-490    242554   296793     539347    204670   288696     493366     262623    302933
5     300-390    113568   133473     247041     82468   131025     213493     146106    144381
6     200-290     30728    29154      59882     18788    26562      45350      32500     24933
Variables not shown: write_total (int)
```

```
sat %>%
  #select() all columns that do NOT contain the word "total"
  select(-contains("total")) %>%
  #gather() all columns EXCEPT score_range, using key = part_sex and value = count.
  gather(part_sex,count , -score_range) %>%
  #separate() part_sex into two separate variables (columns), called "part" and "sex", respectively
  separate(part_sex, into=c("part","sex")) %>%
  print
  
   score_range  part  sex  count
1      700–800  read male  40151
2      600-690  read male 121950
3      500-590  read male 227141
4      400-490  read male 242554
5      300-390  read male 113568
6      200-290  read male  30728
7      700–800  read  fem  38898
8      600-690  read  fem 126084
9      500-590  read  fem 259553
10     400-490  read  fem 296793
11     300-390  read  fem 133473
12     200-290  read  fem  29154
13     700–800  math male  74461
14     600-690  math male 162564
15     500-590  math male 233141
16     400-490  math male 204670
17     300-390  math male  82468
18     200-290  math male  18788
19     700–800  math  fem  46040
20     600-690  math  fem 133954
21     500-590  math  fem 257678
22     400-490  math  fem 288696
23     300-390  math  fem 131025
24     200-290  math  fem  26562
25     700–800 write male  31574
26     600-690 write male 100963
27     500-590 write male 202326
28     400-490 write male 262623
29     300-390 write male 146106
30     200-290 write male  32500
31     700–800 write  fem  39101
32     600-690 write  fem 125368
33     500-590 write  fem 247239
34     400-490 write  fem 302933
35     300-390 write  fem 144381
36     200-290 write  fem  24933
```
```
sat %>%
  #select() all columns that do NOT contain the word "total"
  select(-contains("total")) %>%
  #gather() all columns EXCEPT score_range, using key = part_sex and value = count.
  gather(part_sex,count , -score_range) %>%
  #separate() part_sex into two separate variables (columns), called "part" and "sex", respectively
  separate(part_sex, into=c("part","sex")) %>%
  #Use group_by() (from dplyr) to group the data by part and sex, in that order
  group_by(part, sex) %>%
  mutate(total = sum(count),
         prop = count / total
  ) %>% print
  
   score_range  part  sex  count  total       prop
1      700–800  read male  40151 776092 0.05173485
2      600-690  read male 121950 776092 0.15713343
3      500-590  read male 227141 776092 0.29267278
4      400-490  read male 242554 776092 0.31253253
5      300-390  read male 113568 776092 0.14633317
6      200-290  read male  30728 776092 0.03959324
7      700–800  read  fem  38898 883955 0.04400450
8      600-690  read  fem 126084 883955 0.14263622
9      500-590  read  fem 259553 883955 0.29362694
10     400-490  read  fem 296793 883955 0.33575578
11     300-390  read  fem 133473 883955 0.15099524
12     200-290  read  fem  29154 883955 0.03298132
13     700–800  math male  74461 776092 0.09594352
14     600-690  math male 162564 776092 0.20946486
15     500-590  math male 233141 776092 0.30040382
16     400-490  math male 204670 776092 0.26371873
17     300-390  math male  82468 776092 0.10626060
18     200-290  math male  18788 776092 0.02420847
19     700–800  math  fem  46040 883955 0.05208410
20     600-690  math  fem 133954 883955 0.15153939
21     500-590  math  fem 257678 883955 0.29150579
22     400-490  math  fem 288696 883955 0.32659581
23     300-390  math  fem 131025 883955 0.14822587
24     200-290  math  fem  26562 883955 0.03004904
25     700–800 write male  31574 776092 0.04068332
26     600-690 write male 100963 776092 0.13009154
27     500-590 write male 202326 776092 0.26069847
28     400-490 write male 262623 776092 0.33839158
29     300-390 write male 146106 776092 0.18825861
30     200-290 write male  32500 776092 0.04187648
31     700–800 write  fem  39101 883955 0.04423415
32     600-690 write  fem 125368 883955 0.14182622
33     500-590 write  fem 247239 883955 0.27969636
34     400-490 write  fem 302933 883955 0.34270183
35     300-390 write  fem 144381 883955 0.16333524
36     200-290 write  fem  24933 883955 0.02820619
```



